Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-04 paper.qmd
revised figures for weather

data("WeatherVienna")
# first part of the code is preprocessing
# this is following the vignette from the package
# see https://cran.r-project.org/web/packages/ShapleyOutlier/vignettes/ShapleyOutlier_examples.html
weather_summer <- WeatherVienna %>% dplyr::select(-c(`t`, t_max, t_min, p_max, p_min)) %>%
  drop_na() %>%
  filter(month %in% c("JUN", "JUL", "AUG")) %>%
  filter(year >= 1955) %>%
  group_by(year) %>%
  dplyr::select(-month) %>%
  summarise(across(.cols = everything(), function(x) mean(x)))

X <- weather_summer %>% dplyr::select(-c(num_frost, num_ice, year))
rownames(X) <- weather_summer$year
#> Warning: Setting row names on a tibble is deprecated.
X <- robStandardize(X)

@@

#preferred solution 5 clusters using dunn/dunn2 index
set.seed(150)
X_km <- kmeans(X_outside_scaled, 5)
# making a vector that labels all points
# use 0 to indicate point is not an outlier
cluster_labels_all <- as_tibble(X,
                               rownames = "year") %>%
  left_join(as_tibble(X_km$cluster,
                              rownames = "year")) %>%
  mutate(cluster = if_else(is.na(value), 0, value))
cluster_labels <- factor(cluster_labels_all$cluster,
                         levels = c(1:5, 0))

set.seed(22)
animate_xy(X, 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather,
                               ellmu = mu), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           ellmu = mu,
           axes="off", col = as.factor(cluster_labels),
           palette = "Purple-Yellow",
           center = FALSE, rescale = FALSE)

>>>>

data("WeatherVienna")
# first part of the code is preprocessing
# this is following the vignette from the package
# see https://cran.r-project.org/web/packages/ShapleyOutlier/vignettes/ShapleyOutlier_examples.html
weather_summer <- WeatherVienna |> dplyr::select(-c(`t`, t_max, t_min, p_max, p_min)) |>
  drop_na() |>
  filter(month %in% c("JUN", "JUL", "AUG")) |>
  filter(year >= 1955) |>
  group_by(year) |>
  dplyr::select(-month) |>
  summarise(across(.cols = everything(), function(x) mean(x)))

X <- weather_summer |> dplyr::select(-c(num_frost, num_ice, year))
rownames(X) <- weather_summer$year
#> Warning: Setting row names on a tibble is deprecated.
X <- robStandardize(X)

@@

#preferred solution 5 clusters using dunn/dunn2 index
set.seed(150)
X_km <- kmeans(X_outside_scaled, 5)
# making a vector that labels all points
# use 0 to indicate point is not an outlier
cluster_labels_all <- as_tibble(X,
                               rownames = "year") |>
  left_join(as_tibble(X_km$cluster,
                              rownames = "year")) |>
  mutate(cluster = if_else(is.na(value), 0, value)) |>
  arrange(cluster)
cluster_labels <- factor(cluster_labels_all$cluster,
                         levels = c(1:5, 0))

X_sorted <- cluster_labels_all |>
  select(-year, -cluster, -value) |>
  rename(atmax = avg_t_max,
         atmin = avg_t_min,
         nsumm = num_summer,
         nheat = num_heat,
         sunh = sun_h,
         nclear = num_clear,
         ncloud = num_cloud,
         rh = rel_hum,
         rhmax = rel_hum_max,
         rhmin = rel_hum_min,
         vwind = wind_v,
         vw60 = num_wind_v60,
         vwmax = wind_v_max,
         prec = precp_sum,
         nprec = num_precp_01)

set.seed(22)
animate_xy(as.matrix(X_sorted), 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather,
                               ellmu = mu), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           ellmu = mu, cex=2, 
           axes="bottomleft",
           col = as.factor(cluster_labels),
           palette = "Purple-Yellow",
           center = FALSE, rescale = FALSE,
           ellmarks = FALSE, axislablong = TRUE)

# color codes
# "#80146E" "#7262B2" "#31A0C2" "#5ECBBB" "#B9E5B9"
# "#F5F2D8"
# wait for getting random starting planes for
# better results
set.seed(55)
animate_xy(X_sorted[cluster_labels == 4,], 
           guided_anomaly_tour(anomaly_index(),
                               ellipse=as.matrix(Sigma),
                               ellc = size_weather,
                               ellmu = mu), 
           ellipse=as.matrix(Sigma), ellc = size_weather,
           ellmu = mu,
           axes="bottomleft",
           col = "#5ECBBB", half_range = 7, cex = 2,
           center = FALSE, rescale = FALSE,
           ellmarks = FALSE, axislablong = TRUE)
